---
title: "Getting Started with CyFj11"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Getting Started with CyFj11}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{CyFj11}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CyFj11)
library(flowWorkspace)
library(flowCore)
library(CytoML)
library(ggcyto)
```

## Introduction

CyFj11 is an R package designed to bridge the gap between FlowJo v11 workspaces and the R flowWorkspace ecosystem. It provides **asymmetric format conversion** capabilities:

* **Import**: Full support for parsing FlowJo v11 workspaces (.flowjo files) into R GatingSet objects
* **Export**: Generation of FlowJo v10 workspace files (.wsp) from GatingSet objects with broad compatibility

This asymmetric design reflects the technical realities of developing against proprietary formats without vendor documentation. While FlowJo v11 uses a complex JSON schema with extensive UUID cross-referencing that makes direct export impractical, the v10 XML format follows the documented Gating-ML 2.0 standard, making it feasible to generate valid workspaces.

## Comparison with Other FlowJo Import Packages

Several Bioconductor packages can import FlowJo workspaces. Here's how CyFj11 compares:

| Package | FlowJo Version | Input Format | Output | Key Features |
|---------|---------------|--------------|--------|--------------|
| **CyFj11** | v11 | `.flowjo` (ZIP/JSON) | `GatingSet` | Native v11 support, preserves hierarchy, extracts compensation |
| **CytoML** | v10 | `.wsp` (XML) | `GatingSet` | Gating-ML 2.0 standard, well-established |
| **flowWorkspace** | v10 | `.wsp` (XML) | `GatingSet` | Basic v10 import |
| **flowWorkspaceData** | v10 | `.wsp` (XML) | `GatingSet` | Example datasets |

**Key advantages of CyFj11:**

1. **Native FlowJo v11 support**: CyFj11 is the only Bioconductor package that directly parses FlowJo v11's `.flowjo` format (ZIP archive with JSON)
2. **No intermediate conversion needed**: Unlike CytoML which requires FlowJo v10 XML, CyFj11 reads v11 workspaces directly
3. **Preserves v11-specific features**: Handles compensation matrices, transformations, and population definitions from v11
4. **Export capability**: Can export `GatingSet` objects back to FlowJo v10 `.wsp` format for round-trip workflows

**When to use CyFj11:**
- You have FlowJo v11 workspace files (`.flowjo` extension)
- You want to import workspaces without manually exporting to v10 format in FlowJo
- You need to preserve v11-specific compensation and transformation information

**When to use CytoML:**
- You have FlowJo v10 workspace files (`.wsp` XML format)
- You need Gating-ML 2.0 standard compliance
- You're working with legacy v10 workflows

## Installation

To install CyFj11, ensure you have the required dependencies:

```{r install, eval=FALSE}
# Install from GitHub (or CRAN if available)
# devtools::install_github("baj12/CyFj11")

# Load the package
library(CyFj11)
```

## Basic Workflow

The typical workflow with CyFj11 involves three steps:

1. Import a FlowJo v11 workspace
2. Analyze the data in R
3. Export results back to FlowJo (v10 format)

Let's walk through this process with a simple example.

### Importing a FlowJo v11 Workspace

First, we'll read a FlowJo v11 workspace file:

```{r read-workspace, eval=TRUE}
# Set the path to your FlowJo workspace file
workspace_path <- system.file("extdata", "test.data.flowjo", package = "CyFj11")

# Read the FlowJo v11 workspace
ws <- read_flowjo11_workspace(workspace_path)

# Examine the workspace structure
print(names(ws))
```

The `read_flowjo11_workspace()` function parses the ZIP-based FlowJo v11 format and extracts the JSON data structures. The returned object contains all the information from the workspace, including groups, samples, gates, and populations.

### Converting to GatingSet

Next, we convert the parsed workspace to a GatingSet object, which is the standard format for flow cytometry data in R:

```{r convert-gs, eval=TRUE}
workspace_fcs <- system.file("extdata", "", package = "CyFj11")
# fj11_to_gatingset() returns a named list of GatingSet objects, one per sample
gsList <- fj11_to_gatingset(
  fj11_workspace = ws,
  group_name = 1,
  path = workspace_fcs
)

pretty_print_flowjo(workspace_path)
# Check the number of samples
length(gsList)

# View sample names for the first sample
flowCore::sampleNames(gsList[[1]])

# Work with a single sample directly
gs <- gsList[[1]]

# Or merge all samples into one GatingSet (may overwrite per-sample transform settings)
# gs_merged <- flowWorkspace::merge_list_to_gs(gsList)
```

**Note:** If you don't know the group name you can leave it empty and you will be prompted to choose:

```
1: Acquired Data
2: Compensation Data
3: Experiment Data
```

**Note:** `fj11_to_gatingset()` returns a **list** of `GatingSet` objects, one per sample. This preserves per-sample gate variations (tailored gates) from FlowJo v11. To work with all samples in a single `GatingSet`, use `flowWorkspace::merge_list_to_gs(gsList)`, but be aware this may discard sample-specific transformation settings.

Note that you'll need to provide the path to your FCS files. The `fj11_to_gatingset()` function will attempt to locate the FCS files referenced in the workspace.

### Examining the GatingSet

Once we have a GatingSet, we can examine its structure:

```{r examine-gs, eval=T}
# View the population hierarchy
gs_get_pop_paths(gs)

# Get population statistics
stats <- gs_pop_get_stats(gs)
head(stats)

plot(gs)
autoplot(gs, "cells", bins = 80)
autoplot(gs, "cd8Pos", bins = 80, axis_inverse_trans = TRUE)

gate <- gh_pop_get_gate(gs, "cd8Pos")
gh = gs[[1]][[1]]
gh_get_transformations(gh,channel = "APC-A")


```

### Exporting Back to FlowJo

After performing analysis in R, you can export your results back to FlowJo v10 format:

```{r export, eval=TRUE}
# Export to FlowJo v10 format
export_path <- tempfile(fileext = ".wsp")
export_flowjo10_workspace(gs[[1]], export_path)

# Verify the export was successful
file.exists(export_path)
```
Note:

warnings about invalid transformations are normal, as we don't have any transformations set in FlowJo 11 and the gatingset.

The exported workspace can be opened in FlowJo v11, which maintains backward compatibility with v10 workspaces.

## Next Steps

For more detailed information about compensation handling, see `vignette("compensation-workflow")`.

For exporting GatingSet objects back to FlowJo v10 format, see `?export_flowjo10_workspace`.

For function-specific documentation, use the help system:
```{r help, eval=FALSE}
?read_flowjo11_workspace
?fj11_to_gatingset
?export_flowjo10_workspace
```

```{r cytoml-read-xml, eval=FALSE}
library(CytoML)
ws <- CytoML::open_flowjo_xml(export_path)
gs_back <- CytoML::flowjo_to_gatingset(ws, name = "All Samples",
                                            path = workspace_fcs)

# Get population statistics
stats <- gs_pop_get_stats(gs_back)
head(stats)

plot(gs_back)
ggcyto::autoplot(gs_back, "cells", bins = 80)
autoplot(gs_back, "cd8Pos", bins = 80)


```

```{r sessioninfo}
sessionInfo()
```
